---
title: Was the balanced stratified coverage sampling not well spatially distributed,
  or was it inadequately implemented?
author: "Abdelkrim Bouasria"
date: "2025-01-06"
output:
  html_document: default
---

## Introduction

This repository contains a reproducible notebook accompanying the comment on [Asrat et al. (2024)](https://doi.org/10.1016/j.geoderma.2024.117116), published in *Geoderma*. The comment is entitled *Was the balanced stratified coverage sampling not well spatially distributed, or was it inadequately implemented?* Please refer to the comment for further details ([Bouasria, 2025]())

In this notebook, I implement the sampling design described in Asrat et al. (2024) using the details provided in the main paper and the supplementary material. I will demonstrate the preparation of the covariates, followed by the execution of ten sampling iterations to evaluate whether the results exhibit clustering in any iteration.

## Covariates

Using the same covariates and following nearly identical steps, we performed ten sampling iterations, each selecting 599 samples, to evaluate whether the distribution exhibited consistent trends. We focused on the primary wheat-growing regions and applied the cropland mask (GFSAD30AFCE; [Xiong et al., 2017](https://doi.org/10.5067/MEaSUREs/GFSAD/GFSAD30AFCE.001)) at a 30-meter resolution to exclude other land use areas. The FAO soil vector map was rasterized to match the cropland raster grid, and the remaining covariates were aggregated in accordance with the methodology described by Asrat et al. (2024), then resampled to a spatial resolution of 30 meters.

### Definition of wheat-growing areas in Morocco

The soil sampling locations were selected to capture soil variability and distribution across Morocco's rainfed wheat-growing regions, with a focus on prioritizing soil types relevant to wheat cultivation. We referred to the official statistics of Morocco (CGDA, 2020; page 96) to select 6 out of the 10 wheat-producing regions, which encompass the primary rainfed cropland areas.

The preparation of covariates was performed using the `terra` package.

```{r warning=FALSE}
library(terra)
```

We sourced cropland data from the GFSAD30AFCE dataset ([Xiong et al., 2017](https://doi.org/10.5067/MEaSUREs/GFSAD/GFSAD30AFCE.001)). Three tiles were combined into a single VRT file, where we assigned `NA` values to non-cropland classes. We then cropped this file to align with the wheat-growing regions mask.

```{r eval=FALSE, include=TRUE}
cropland = rast("./_RAW_Data/CROPLAND/croplands.vrt")
cropland[cropland==0] = NA
cropland[cropland==1] = NA

rv_mask= vect("./vect/Morocco_regions_mask.shp")

crop(cropland,rv_mask, mask=TRUE, filename="./covariates/croplands.tif",
            datatype="INT1U", gdal=c("COMPRESS=DEFLATE"))

```

```{r}
rv= vect("./vect/Morocco_regions.shp")
cropland = rast("./covariates/croplands.tif")
plot(cropland, pax=list(retro=TRUE, side=c(1:4)))
polys(rv)
text(rv,"NAME_1", halo=TRUE)
```

### FAO Soil map

Soil data was sourced from the Digital Soil Map of the World (DSMW), which is based on the FAO-UNESCO Soil Map of the World. The DSMW, digitized at a 1:5,000,000 scale, uses a Geographic projection (latitude-longitude). Further details about the DSMW can be found on the [FAO website](http://www.fao.org/soils-portal/soil-survey/soil-maps-and-databases/faounesco-soil-map-of-the-world/en/). The vector dataset is available for download via the [FAO data catalog](https://data.apps.fao.org/map/catalog/srv/eng/catalog.search#/metadata/446ed430-8383-11db-b9b2-000d939bc5d8) or directly as a ZIP file: [DSMW.zip](https://storage.googleapis.com/fao-maps-catalog-data/uuid/446ed430-8383-11db-b9b2-000d939bc5d8/resources/DSMW.zip).  

The vector map was clipped using the wheat-growing areas vector mask.  

```{r}
rv_mask= vect("./vect/Morocco_regions_mask.shp")
soil = crop(vect("./_RAW_Data/FAO_SOIL/DSMW/DSMW.shp"),rv_mask)
plot(soil,"DOMSOI", ext=ext(rv))
```
The vector file was then rasterized to align with the cropland raster grid cells.  

```{r eval=FALSE, include=TRUE}
mask(rasterize(soil,cropland, field="DOMSOI"),
                cropland, filename="./covariates/fao_soil.tif",
                     datatype="INT1U", gdal=c("COMPRESS=DEFLATE"))
```

The resulting raster grid contained 23 different soil types.  

```{r}
rst_soil = rast("./covariates/fao_soil.tif")
plot(rst_soil, type="classes", col=hcl.colors(23,"Roma"))
polys(rv)
```

### Topographic covariates: elevation and slope

We used the QGIS SRTM-Downloader plugin to download 25 SRTM DEM tiles from the NASA server. These tiles were then merged into a single VRT file for further processing.

```{r}
dem = rast("./_RAW_Data/DEM/dem.vrt")
plot(dem, col= rev(hcl.colors(10, "Greens 3")))
polys(rv)

```
Elevation and slope data were prepared using the `terra` package, and the results were resampled to align with the cropland raster grid cells.

```{r eval=FALSE, include=TRUE}
resample(crop(dem,rv_mask, mask=TRUE),
         cropland, method="bilinear", threads=TRUE,
         filename="./covariates/elevation.tif",
         gdal=c("COMPRESS=DEFLATE"),overwrite=TRUE)

slope = resample(crop(terrain(dem, v="slope"),rv_mask, mask=TRUE),
                 cropland, method="bilinear", threads=TRUE,
                 filename="./covariates/slope.tif",
                 gdal=c("COMPRESS=DEFLATE"),overwrite=TRUE)
```


```{r fig.height=3.5, fig.width=8}
elevation = rast("./covariates/elevation.tif")
slope = rast("./covariates/slope.tif")

plot(c(elevation,slope), main=c("Elevation","Slope"),
     fun = \() polys(rv))
```

### Rainfall: Five years annual average

The five-year mean (2015–2019) annual rainfall data for the African continent, with a 5 km spatial resolution, was obtained from CHIRPS and is accessible at [https://data.chc.ucsb.edu/products/CHIRPS-2.0/africa_dekad/tifs/](https://data.chc.ucsb.edu/products/CHIRPS-2.0/africa_dekad/tifs/). The CHIRPS decadal rainfall data for Africa were decompressed and cropped to align with the study area.  


```{r eval=FALSE, include=TRUE}
library(R.utils)
# Crop for the study area extent
dir_in  = "./_RAW_Data/CHIRPS/"
dir_out = "./_RAW_Data/temporary/"
gz_files =  list.files(dir_in, pattern = ".gz$")


for(gz in gz_files){
  tif_files = rast(gunzip(paste0(dir_in,gz)))
  terra::crop(tif_files,rv_mask, mask=TRUE, filename=paste0(dir_out,gsub(".gz","",gz)))
  cat(gz,"\n")
}

```

We calculated the cumulative rainfall for each year and then derived the five-year annual average rainfall. Finally, the average rainfall data were resampled to match the cropland layer.

```{r eval=FALSE, include=TRUE}

# calculate the cumulative annual rainfall
rst_files =  list.files(dir_out, pattern = ".tif$")
years =  sapply(gsub(".tif$","",gsub("chirps-v2.0.","",rst_files)),
              \(x) strsplit(x,"\\.")[[1]][1])

df_files = data.frame(files=rst_files,years=years )

annual_rainfall = NULL

for(year in unique(df_files$years)){
  df = df_files[df_files$years==year,]
  rainfall_year = sum(rast(paste0(dir_out,df$files)))
  
  if(is.null(annual_rainfall)){
  annual_rainfall = rainfall_year
    }else{
  annual_rainfall = c(annual_rainfall, rainfall_year)
    }
}

# calculate the five years annual avg
annual_rainfall_avg = mean(annual_rainfall)

# resemple to cropland raster
resample(annual_rainfall_avg,
                    cropland, method="bilinear", threads=TRUE,
                    filename="./covariates/rainfall.tif",
                    gdal=c("COMPRESS=DEFLATE"),overwrite=TRUE)

```


```{r fig.width=6}
rainfall = rast("./covariates/rainfall.tif")
plot(rainfall)
polys(rv)
# 
```

### Temperature: Five years annual minimum and maximum temperature

Five years (2015–2019) of annual minimum and maximum temperature data with a 50 km spatial resolution were downloaded as a '.NC' file from the [NOAA CPC Global Temperature Dataset](https://downloads.psl.noaa.gov/Datasets/cpc_global_temp/). The rasters were rotated to align with the correct world map using the `terra` package. The average temperature for each year was then calculated from the monthly maximum and minimum temperatures, followed by calculating the five-year average. Finally, the result was resampled to align with the cropland layer.


```{r eval=FALSE, include=TRUE}

rst_files =  list.files("./Temp_NOAA/", pattern = ".nc$")
temp1 = rast(paste0("./Temp_NOAA/",rst_files[1]))
plot(temp1[[1]])
x <- rotate(temp1) 
plot(x[[1]])

rv1 = buffer(rv,5000)

x1 = crop(x,rv1, mask=TRUE)
plot(x1[[1]])
polys(rv)
polys(rv1)

for(rst in rst_files){
  terra::crop(rotate(rast(paste0("./Temp_NOAA/",rst))),rv1, mask=TRUE, filename=paste0("./Temp_NOAA_MA/",rst))
  cat(rst,"\n")
}

# calculate annual AVG of Tmin and Tmax
rst_files =  list.files("./Temp_NOAA_MA/", pattern = ".nc$")

years =  sapply(gsub(".nc$","",rst_files),
                \(x) strsplit(x,"\\.")[[1]][2])
temp =  sapply(gsub(".nc$","",rst_files),
                \(x) strsplit(x,"\\.")[[1]][1])

df_files = data.frame(files=rst_files,temp=temp,years=years )

df_tmax = df_files[df_files$temp=="tmax",]
df_tmin = df_files[df_files$temp=="tmin",]

annual_tmax = lapply(df_tmax$years, function(year){
  mean(rast(paste0("./Temp_NOAA_MA/",df_tmax[df_tmax$years==year,]$files)))
})

annual_tmax_avg = mean(rast(annual_tmax))

annual_tmin = lapply(df_tmin$years, function(year){
  mean(rast(paste0("./Temp_NOAA_MA/",df_tmin[df_tmax$years==year,]$files)))
})

annual_tmin_avg = mean(rast(annual_tmin))


tmax = resample(annual_tmax_avg,
                cropland, method="bilinear", threads=TRUE,
                filename ="./covariates/tmax.tif", 
                gdal=c("COMPRESS=DEFLATE"),overwrite=TRUE)
tmin = resample(annual_tmin_avg,
                cropland, method="bilinear", threads=TRUE,
                filename ="./covariates/tmin.tif", 
                gdal=c("COMPRESS=DEFLATE"),overwrite=TRUE)


```


```{r fig.height=3.5, fig.width=8}
temp = c(rast("./covariates/tmax.tif"),rast("./covariates/tmin.tif"))
plot(temp, fun= \() polys(rv))
```



To optimize the raster for download on GitHub, which limits file size to 100 MB, we tiled the covariates and merged them into a VRT file.

```{r}
rv_mask= vect("./vect/Morocco_regions_mask.shp")

r_files = list.files("./covariates/", pattern = ".tif$", full.names = T)[-1]
cov = rast(lapply(r_files, \(r) rast(r)))
my_crs = crs(cov)

grid = getTileExtents(cov, c(1024,1024)*2, extend=TRUE, buffer=2)

vgrid = NULL

for(i in 1:nrow(grid)){
  v = as.polygons(ext(grid[i,]),crs=my_crs)
  if(is.null(vgrid)){
    vgrid = v
  }else{ vgrid= rbind(vgrid,v)}
}

vgrid = crop(vgrid, rv_mask)

plot(cov[[1]], fun=function() polys(rv))
polys(vgrid)

```

We generated 76 tiles, each with a smaller size.

```{r eval=FALSE, include=TRUE}
cov <- makeTiles(cov, vgrid, na.rm = TRUE, 
                      filename = "./covariates2/cov_30_m_.tif",
                      gdal=c("COMPRESS=DEFLATE"))

```


## Sampling design test

The authors utilized the `lcubestratified` function from the `BalancedSampling` R package, which implements stratified doubly balanced sampling with pooling of landing phases, utilizing the fast flight Cube method. This function requires seven input arguments, but only four are specified in the paper. These include: (1) the inclusion probabilities (`prob`), which were set to be equal for all points; (2) the spreading parameter (`Xpread`), representing the geographical coordinates across the study area; (3) the balancing auxiliary variables (`Xbal`), which include rainfall, temperature, slope, and elevation; and (4) the stratification parameter (`strata`), representing the FAO soil types . The settings for the remaining arguments were not reported, and therefore, it is assumed that they were left at their default values.
```{r}
r_files = list.files("./covariates2/", pattern = ".tif$", full.names = TRUE)
cov = vrt(r_files, filename= "./covariates2/cov_30_m.vrt", overwrite=TRUE)

names(cov) = c("elevation", "soil", "rainfall", "slope", "tmax", "tmin")
plot(cov, fun=function() polys(rv))
```


This code will take a long time to run and requires more than 120 GB of memory. It cannot be executed on a laptop or ordinary desktop computers.

```{r eval=FALSE, include=TRUE}

library(BalancedSampling)

rv= vect("./vect/Morocco_regions_mask.shp")
cov = as.data.frame(cov, xy=T,na.rm=TRUE)

saveRDS(cov,"./covariates/cov_30m_df.rds")

N = nrow(cov)
n = 599
sprob = rep(n/N, N)

xspr = cov[,c("x","y")]
xbal  = cov[,c("elevation", "slope", "rainfall", "tmax", "tmin")]
strata = cov[,"soil"]

iter = 10
set.seed(1)
seeds <- runif(iter,1,100000)|> round(0)

samples = list()

for(i in 1:iter){
  set.seed(seeds[i])
  samples[[i]] = lcubestratified(prob = sprob,
                                 Xspread = xspr,
                                 Xbal = xbal,
                                 integerStrata = strata)
  cat("iter:",i,"\n")
}


saveRDS(samples,"./samples/samples.rds")
saveRDS(xspr,"./samples/cov_xy.rds")

```

After generating the samples for each iteration, we combined them into a single vector point file.

```{r eval=FALSE, include=TRUE}
samples = readRDS("./samples/samples.rds")
xspr = readRDS("./samples/cov_xy.rds")
crs_xy = crs(rast("./cov/cov_30m.tif"))

samples_xy = lapply(samples,\(x) xspr[x,])
for(i in 1:length(samples_xy)){
  samples_xy[[i]]=cbind(samples_xy[[i]], iter = i)
}
samples_c = do.call(rbind,samples_xy)

saveRDS(samples_c,"./samples_c.rds")

samples_vect = vect(samples_c, geom=c("x", "y"), crs=crs_xy)
writeVector(samples_vect,"./samples/samples_vect.gpkg")

```

The results of the sampling design are shown below.  

```{r}
rv= vect("./vect/Morocco_regions_mask.shp")
samples_vect = vect("./samples/samples_vect.gpkg")

plot(samples_vect,"iter", cex=0.5,mar=c(1,1,1,1),
     plg=list(x="topleft",title="Iteration", title.cex=1.2, cex=1.2), 
     pax=list(side=1:4, retro=TRUE,cex.axis=1.2))
polys(rv)
north("top")
sbar(xy="bottomright", type="bar")

```

The results of the sampling design for each iteration are shown separately below.  

```{r fig.height=10, fig.width=8}
par(mfrow = c(4,3))

for(i in 1:10){
  plot(samples_vect[samples_vect$iter==i],cex=0.5, col="blue",ext=ext(rv),
       main=paste0("sampling iteration: ",i), mar=c(1,1,1.8,0.5))
  polys(rv)
}
```


